---
title: "Imagine -- An Exploratory Analysis"
output: 
  pdf_document:
    keep_tex: true
    citation_package: natbib
biblio-style: apalike
tables: true
header-includes: 
  - \usepackage{booktabs}
  - \usepackage{threeparttable}
  - \usepackage[utf8]{inputenc}
  - \usepackage[T1]{fontenc}
  - \usepackage{amsmath}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{pdflscape}
  - \usepackage{float} #use the 'float' package
  - \floatplacement{figure}{H} #make every figure with caption = h
  - \usepackage{subcaption}
date: "`r Sys.Date()`"
bibliography: hb.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
options(knitr.kable.NA = '')
```

### Before we start

The data used in this report is the same you used for your analysis. I used the mean ratings for each dimension (factivity, intentionality, pictoriality) and scaled them. In the upcoming weeks, I will also adapt the scripts as to use the ratings of both coders to account for between-coder variance. I just didn't have the time to do everything. Since manual sense annotations are not yet available, I wrote the pipeline based on a clustering of the the data on the three provided annotation dimensions. We can later just replace them with the actual annotations.

### Pre-analysis

```{r}
library(readxl)
library(dplyr)
library(mclust)
library(ggplot2)
library(reshape2)
library(scatterplot3d)
library(tidyr)
library(glmnet)
library(nnet)
library(purrr)
library(scales)
library(kableExtra)

rm(list = ls())
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```


Table \ref{tab:pairwisePearsonDimensions}.

```{r}
# Load data
df <- read_xlsx("../../input/corpus_annotated.xlsx", sheet = 1)
names(df) <- gsub("\\smeans", "", tolower(names(df)))

# Drop excluded snippets and those with too little information
df <- df %>% mutate(across(ends_with("ity"), ~ as.numeric(.))) # will introduce NAs for X and E
df <- na.omit(df)

# Select only the annotation dimensions
annot <- df %>%
  select(intentionality, factivity, pictoriality)

# Pairwise correlations
pairwiseCor <- round(cor(annot, use = "pairwise.complete.obs"), 3)
pairwiseCor[lower.tri(pairwiseCor, diag = F)] <- NA

# Table
kable(pairwiseCor, booktabs = TRUE, caption = "Pairwise Pearson Correlation", label = "pairwisePearsonDimensions", format = "latex") %>% kable_styling(latex_options = "HOLD_position")
```

```{r}
# Standardize annotation space
annot_z <- as.data.frame(scale(annot))
```

Figure \ref{fig:gmmBIC}

```{r fig.width=6, fig.height=4, fig.pos="h", fig.cap="BIC for Gaussian mixture models.\\label{fig:gmmBIC}"}
par(cex.axis = 0.75, cex.lab  = 0.75, mgp = c(2, 0.7, 0))
## Clustering
# 1- vs 2- vs 3-component solution
gmm_res <- Mclust(annot_z, G = 1:3)

# BIC values
plot(gmm_res, what = "BIC",
     legendArgs = list(x = "topleft", cex = 0.5)
     )
```


Figure \ref{fig:pcaGMM}.

```{r fig.width=4, fig.height=3, fig.pos="h", fig.cap="PCA of annotation space with latent sense clusters.\\label{fig:pcaGMM}"}
# option with three clusters is preferred, but we use a two-cluster solution for the mock up

# 2-component solution
gmm_res <- Mclust(annot_z, G = 2)

# Cluster assignments
annot_z$sense <- gmm_res$classification

# Posterior probabilities (membership confidence)
annot_z$posterior_max <- apply(gmm_res$z, 1, max)

# PCA to check for 2D cluster separation
pca_res <- prcomp(annot_z, center = FALSE, scale. = FALSE)

# Variance explained
pca_summary <- summary(pca_res)

# Combine clustering with PCA
pca_res <- bind_cols(pca_res$x, annot_z)

ggplot(pca_res, aes(PC1, PC2, color = as.factor(paste0("sense ", sense)))) +
  geom_point(alpha = 0.2) +
  labs(title = "",
       x = paste0("PC1 (", round(pca_summary$importance[2, 1]*100, 2), "%)"),
       y = paste0("PC2 (", round(pca_summary$importance[2, 2]*100, 2), "%)"),
       color = "Cluster") +
  theme_classic()
```


```{r out.width='.32\\linewidth', fig.width=3, fig.height=3, fig.show='hold', fig.align='center', fig.cap="Predicted sense probability based on GLM with ridge regularization for 2-way interactions. The third dimension is held constant at the median value.The lighter the color, the higher the probability.\\label{fig:xEffects}"}
### GLM with ridge regularization
# Fit multinomial regression with 2-way interactions as IV L2 (ridge) regularization for stability
y <- as.factor(paste0("sense ", annot_z$sense))
x_int <- model.matrix(~ (intentionality + factivity + pictoriality)^2, data = annot_z)[,-1]
cv_fit_int <- cv.glmnet(x_int, y, family = "multinomial", alpha = 0, type.measure = "class")
best_lambda <- cv_fit_int$lambda.min

coefs <- coef(cv_fit_int, s = best_lambda)

# Convert to data frame for inspection
coefs_df <- do.call(rbind, lapply(names(coefs), function(cluster){
  data.frame(cluster = cluster,
             term = rownames(coefs[[cluster]]),
             coef = as.numeric(coefs[[cluster]]))
}))

# Plot interactions
vars <- c("intentionality", "factivity", "pictoriality")
pairs <- combn(vars, 2, simplify = FALSE)

plots <- list()

for (vp in pairs) {
  x_var <- vp[1]
  y_var <- vp[2]
  fix_var <- setdiff(vars, vp)
  
  # Build grid manually
  grid <- expand.grid(
    seq(min(annot_z[[x_var]]), max(annot_z[[x_var]]), length.out = 50),
    seq(min(annot_z[[y_var]]), max(annot_z[[y_var]]), length.out = 50)
  )
  colnames(grid) <- c(x_var, y_var)
  
  # Add fixed variable at its median
  grid[[fix_var]] <- median(annot_z[[fix_var]])
  
  # Model matrix
  grid_mm <- model.matrix(~ (intentionality + factivity + pictoriality)^2, data = grid)[,-1]
  
  # Predicted probabilities
  probs <- predict(cv_fit_int, newx = grid_mm, type = "response", s = best_lambda)
  probs <- as.data.frame(probs[,,1])
  colnames(probs) <- levels(y)
  grid <- cbind(grid, probs)
  
  # Long format
  grid_long <- pivot_longer(grid, cols = colnames(probs), names_to = "cluster", values_to = "prob") %>%
    group_by(cluster) %>%
    mutate(prob_scaled = rescale(prob)) %>%
    ungroup()
  
  # Heatmap plot
  p <- ggplot(grid_long, aes_string(x = x_var, y = y_var, fill = "prob_scaled")) +
    geom_raster() +
    facet_wrap(~ cluster, ncol = 1) +
    scale_fill_viridis_c(option = "magma") +
    labs(x = paste0(x_var, " (z-score)"),
         y = paste0(y_var, " (z-score)"),
         fill = "Predicted probability",
         #title = paste("Cluster probabilities:", x_var, "vs", y_var)
         ) +
    scale_y_continuous(expand = c(0,0)) +
    scale_x_continuous(expand = c(0,0)) +
    theme_light() +
    theme(
      panel.grid = element_blank(),
      legend.position = "none"
    )

  plots[[paste(x_var, y_var, sep = "_")]] <- p
}

plots$intentionality_factivity
plots$intentionality_pictoriality
plots$factivity_pictoriality
```
```{r fig.width=5, fig.height=2, fig.pos="h", fig.cap="Bootstrapped Coefficients (n = 500) with 95% CI.\\label{fig:bootstrapping}"}
# Robustness check via bootstrapping
n_boot <- 500
n_vars <- ncol(x_int)
n_clusters <- length(levels(y))

# Arrays to store results
coefs_boot <- array(NA, dim = c(n_boot, n_vars, n_clusters))
selected_boot <- array(0, dim = c(n_boot, n_vars, n_clusters))  # 1 if nonzero

set.seed(1847)
for(i in 1:n_boot){
  # resample rows with replacement
  rows <- sample(1:nrow(x_int), replace = TRUE)
  
  # fit multinomial glmnet
  fit <- glmnet(x_int[rows,], y[rows], family="multinomial", alpha=0)
  
  # extract coefficients at best lambda
  coefs_list <- coef(fit, s = best_lambda)
  
  for(k in 1:n_clusters){
    # coefs_list[[k]] includes intercept as row 1, remove intercept if desired
    coefs_k <- as.numeric(coefs_list[[k]][-1, ]) # drop intercept
    coefs_boot[i, , k] <- coefs_k
    
    # mark which coefficients are non-zero
    selected_boot[i, , k] <- as.numeric(coefs_k != 0)
  }
}

# Compute bootstrap CI
ci_lower <- apply(coefs_boot, c(2,3), function(x) quantile(x, 0.025, na.rm=TRUE))
ci_upper <- apply(coefs_boot, c(2,3), function(x) quantile(x, 0.975, na.rm=TRUE))
coef_median <- apply(coefs_boot, c(2,3), median, na.rm=TRUE)

# Compute stability (selection frequency)
stability <- apply(selected_boot, c(2,3), mean, na.rm=TRUE)

# Convert to data.frame for reporting
rename_terms <- function(terms) {
  parts <- strsplit(terms, ":", fixed = TRUE)
  
  renamed <- vapply(parts, function(p) {
    paste(toupper(substr(p, 1, 1)), collapse = ":")
  }, character(1))
  
  renamed
}

coef_df <- expand.grid(
  term = rename_terms(colnames(x_int)),
  cluster = as.factor(paste0("sense ", 1:n_clusters))
) %>%
  mutate(
    median = as.vector(coef_median),
    ci_lower = as.vector(ci_lower),
    ci_upper = as.vector(ci_upper),
    stability = as.vector(stability)
  )

ggplot(coef_df, aes(x = term, y = median, color = cluster)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.6),
                width = 0.2, size = 0.25) +
  geom_point(position = position_dodge(width = 0.6), size = 0.25) +
  coord_fixed() +
  labs(
    color = "Cluster",
    x = "Term",
    y = "Median Coefficients"
  ) +
  theme_light()

```

# Embedding analysis

## Centroids

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{../../output/plots/overlap_density.png}
    \caption{Caption}
    \label{fig:placeholder}
\end{figure}

\begin{figure}
\centering
    \begin{subfigure}[b]{0.6\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../../output/plots/kde1D_embeddings.png}
    \caption{\label{fig:image1}}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.38\textwidth}
    \centering
    \includegraphics[width=0.9\textwidth]{../../output/plots/kde2D_embeddings.png}
    \caption{\label{fig:image2}}
    \end{subfigure}
\end{figure}

## Embedding similarity vs sense similarity

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{../../output/plots/embedding_vs_sense_similarity.png}
    \caption{Caption}
    \label{fig:placeholder}
\end{figure}

## Senses under embedding projection

To examine how sentence embeddings capture the three annotated semantic dimensions (*intentionality*, *factivity*, and *pictoriality*), we trained simple linear regression models predicting each z-scaled dimension from the sentence embeddings. The regression coefficients define a direction in embedding space along which the dimension increases. By projecting each sentence embedding onto these directions, we obtain an embedding-based estimate for that dimension. 
This procedure was repeated for all three dimensions, resulting in a 3D projection of each sentence in the space spanned by the semantic dimensions.

Let $\mathbf{X} \in \mathbb{R}^{n \times d}$ denote the matrix of sentence embeddings, where $n$ is the number of sentences and $d$ the embedding dimension. Let $\mathbf{y}^{(k)} \in \mathbb{R}^n$ be the vector of z-scaled ratings for semantic dimension $k \in \{\text{intentionality, factivity, pictoriality}\}$. 

We fit a simple linear regression model for each dimension:

\[
\mathbf{y}^{(k)} = \mathbf{X} \boldsymbol{\beta}^{(k)} + \boldsymbol{\epsilon}^{(k)},
\]

where $\boldsymbol{\beta}^{(k)} \in \mathbb{R}^d$ is the vector of regression coefficients and $\boldsymbol{\epsilon}^{(k)}$ is the residual vector. The estimated coefficients $\hat{\boldsymbol{\beta}}^{(k)}$ define a direction in embedding space corresponding to increases in dimension $k$. 

The projection of each sentence embedding $\mathbf{x}_i$ onto this direction provides an embedding-based estimate of the semantic dimension:

\[
\hat{y}^{(k)}_i = \mathbf{x}_i^\top \hat{\boldsymbol{\beta}}^{(k)}, \quad i = 1, \dots, n.
\]

By computing these projections for all three dimensions, each sentence is represented in a 3D space spanned by the semantic directions, allowing visualization and analysis of how embeddings capture the annotated semantic structure.

Figure \ref{fig:3Dprojection} shows 3D projection of sentence embeddings onto the directions corresponding to the annotated semantic dimensions (intentionality, factivity, and pictoriality). It reveals that the sense clusters are well-separated in this space. This indicates that the latent embedding space encodes information aligned with the annotated dimensions. In other words, the directions learned via linear regression effectively capture the semantic structure underlying the sense distinctions. The clear separation suggests that even a simple linear mapping from embeddings to semantic dimensions is sufficient to discriminate the senses, highlighting that the embeddings intrinsically encode patterns that correspond to the interpretive distinctions annotated in the dataset.

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{../../output/plots/embedding_projection_3D.png}
    \caption{3D embedding projections. Dimensions have been rescaled to $z$-scores.}
    \label{fig:3Dprojection}
\end{figure}

Figure \ref{fig:2Dprojection} depicts 2-dimensional for all combinations of dimensions. For projections for factivity and intentionality (Fig. \ref{fig:2D_IF}), sense 1 was predominantly associated with high factivity and low intentionality, whereas sense 2 showed the opposite pattern. Both distributions exhibited slight bimodality. In the pictoriality–intentionality projection (Fig. \ref{fig:2D_IP}), pictoriality contributed minimally, with sense clusters largely segregated along the intentionality axis. In contrast, the pictoriality–factivity projection (Fig. \ref{fig:2D_FP}) revealed substantial overlap between the two senses, indicating that the senses are not well distinguished along this dimension pair. These results suggest that the distinctions between the senses are primarily encoded along the intentionality and factivity dimensions, whereas pictoriality alone does not reliably differentiate sense membership.

\begin{figure}
\centering
    \caption{2D embedding projections. Dimensions have been rescaled to $z$-scores.}
    \label{fig:2Dprojection}
    \begin{subfigure}[b]{0.30\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../../output/plots/embedding_projection_intentionality_z_factivity_z_2D_core.png}
    \caption{\label{fig:2D_IF}}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.30\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../../output/plots/embedding_projection_intentionality_z_pictoriality_z_2D_core.png}
    \caption{\label{fig:2D_IP}}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.30\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../../output/plots/embedding_projection_factivity_z_pictoriality_z_2D_core.png}
    \caption{\label{fig:2D_FP}}
    \end{subfigure}
\end{figure}

